Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-10-24 paper-new/intro.tex
small tweaks

\emph{Gradually-typed languages} \cite{siek-taha06, tobin-hochstadt06} seek to
resolve this tension by allowing for both static and dynamic typing disciplines
to be used in the same codebase. These languages support smooth interoperability
between statically-typed and dynamically-typed styles, allowing the programmer to
begin with fully dynamically-typed code and \emph{gradually} migrate portions of the
codebase to a statically typed style without needing to rewrite the project in a
completely different language.

@@

Additionally, gradually typed languages should offer some of the
benefits of static typing. While classical type soundness, that
well-typed programs are free from runtime errors, is not compatible
with runtime type errors, we can instead focus on \emph{type-based
reasoning}. For instance, while dynamically typed $\lambda$ calculi
only satisfy $\beta$ equality for their type formers, statically typed
$\lambda$ calculi additionally satisfy type-dependent $\eta$
properties that ensure that functions are determined by their behavior
under application and that pattern matching on data types
is safe and exhaustive.

@@

The only prior denotational semantics for gradual typing was given by
New and Licata and is based on a classical Scott-style \emph{domain
theory} \cite{new-licata18}. The fundamental idea is to equip
$\omega$-CPOs with an additional ``error ordering'' $\ltdyn$ which
models the graduality ordering, and for casts to arise from
\emph{embedding-projection pairs}. Then the graduality property
follows as long as all language constructs can be interpreted using
constructions that are monotone with respect to the error ordering.
%
This framework has the benefit of being compositional, and was
expressive enough to be extended to model dependently typed gradual
typing \cite{gradualizing-cic}.
%
\max{we should make clear that we don't yet model these perverse recursive features}
\eric{In the contributions section, I explain that we don't yet model these features}
However, an approach based on classical domain theory has fundamental
limitations: domain theory is incapable of modeling certain perversely
recursive features of programming languages such as dynamic type tag
generation and higher-order references, which are commonplace in
real-world gradually typed systems as well as gradual calculi.
%
Our long-term goal is to develop a semantics that will scale up to these
advanced features, and so we begin with this work by departing from
classical domain theory.

The main denotational alternative to classical domain theory that can
successfully model these advanced features is \emph{guarded} domain
theory, which we adopt in this work. While classical domain theory is
based on modeling types as ordered sets with certain joins, guarded
domain theory is based on an entirely different foundations, sometimes
(ultra)metric spaces but more commonly as ``step-indexed sets'', i.e.,
objects in the \emph{topos of trees}
\cite{birkedal-mogelberg-schwinghammer-stovring2011}.  Such an object
consists of a family $\{X_n\}_{n \in \mathbb{N}}$ of sets along with
restriction functions $r_n : X_{n+1} \to X_n$ for all $n$.  (in
category theoretic terminology, these are presheaves on the poset of
natural numbers.)  We think of a $\mathbb{N}$-indexed set as an infinite
sequence of increasingly precise approximations to the true type being
modeled.
%
Key to guarded domain theory is that there is an operator
$\triangleright$ on step-indexed sets called ``later''. In terms of
sequences of approximations, the later operator delays the
approximation by one step. Then the crucial axiom of guarded domain
theory is that any guarded domain equation $X \cong F(\triangleright
X)$ has a unique solution. This allows guarded domain theory to model
essentially \emph{any} recursive concept, with the caveat that the
recursion is \emph{guarded} by a use of the later operator.

% The theorems we want are about the whole sequence of approximations
% whereas working with presheaves is about each finite approximation
% The statement of graduality is in terms of global elements
\eric{This next paragraph could be moved to the section on adequacy, but since
the example is in the analytic setting it also makes sense to keep it here. I'm
not sure which option makes more sense.} While the definitions in guarded domain
theory are constructed as sequences of approximations, results about semantics
of programs, e.g., graduality, are actually statements concerning entire
\emph{sequences} of approximations. As a simple example, consider the set of
programs that may take a computational step or return unit.\footnote{This
example is adapted from \cite{mogelberg-paviotti2016}.} In the topos of trees,
we can model the set of these programs as the indexed family of sets $\{X_n\}_{n
\in \mathbb{N}}$ where for each $n \ge 1$ we define $X_n = \{0,1,\dots,n-1,
\bot\}$. Here, $i$ denotes a program that steps $i$ times and then returns, and
$\bot$ denotes a program that fails to terminate in $n-1$ steps or fewer. For
any fixed $n$, the set $X_n$ fails to distinguish a program that takes $n$ steps
and then terminates from a program that never terminates. On the other hand, if
we define the denotation of a program to be a \emph{sequence} of elements $x_i
\in X_i$ for all $i \in \mathbb{N}$, then the denotation of the diverging
program will be distinct from that of a program that terminates, regardless of
the number of steps it takes.

>>>>

\emph{Gradually typed languages} \cite{siek-taha06, tobin-hochstadt06} seek to
resolve this tension by allowing for both static and dynamic typing disciplines
to be used in the same codebase. These languages support smooth interoperability
between statically-typed and dynamically-typed styles, allowing the programmer to
begin with fully dynamically-typed code and \emph{gradually} migrate portions of the
codebase to a statically typed style without needing to rewrite the project in a
completely different language.

@@

Additionally, gradually typed languages should offer some of the
benefits of static typing. While standard type soundness, that
well-typed programs are free from runtime errors, is not compatible
with runtime type errors, it is possible instead to prove that
gradually typed languages validate \emph{type-based reasoning}. For
instance, while dynamically typed $\lambda$ calculi only satisfy
$\beta$ equality for their type formers, statically typed $\lambda$
calculi additionally satisfy type-dependent $\eta$ properties that
ensure that functions are determined by their behavior under
application and that pattern matching on data types is safe and
exhaustive. A gradually typed calculus that validates these
type-dependent $\eta$ laws then provides some of the type-based
reasoning that dynamic languages lack.

@@

The only prior denotational semantics for gradual typing was given by
New and Licata and is based on a classical Scott-style \emph{domain
theory} \cite{new-licata18}. The fundamental idea is to equip
$\omega$-CPOs with an additional ``error ordering'' $\ltdyn$ which
models the graduality ordering, and for casts to arise from
\emph{embedding-projection pairs}. Then the graduality property
follows as long as all language constructs can be interpreted using
constructions that are monotone with respect to the error ordering.
%
This framework has the benefit of being compositional, and was
expressive enough to be extended to model dependently typed gradual
typing \cite{gradualizing-cic}.
%
However, an approach based on classical domain theory has fundamental
limitations: domain theory is incapable of modeling certain perversely
recursive features of programming languages such as dynamic type tag
generation and higher-order references, which are commonplace in
real-world gradually typed systems as well as gradual calculi.
%
Our long-term goal is to develop a denotational approach that can
scale up to these advanced features, and so we must abandon classical
domain theory as the foundation in order to make progress. In this
work, we provide first steps towards this goal by adapting work on a
simple gradually typed lambda calculus to \emph{guarded domain
theory}. Our goal is for this model to scale to dynamic type tag
generation and higher-order references in future work.

\emph{Guarded} domain theory is the main denotational alternative to
classical domain theory that can successfully model these advanced
features. While classical domain theory is based on modeling types as
ordered sets with certain joins, guarded domain theory is based on an
entirely different foundations, sometimes (ultra)metric spaces but
more commonly as ``step-indexed sets'', i.e., objects in the
\emph{topos of trees}
\cite{birkedal-mogelberg-schwinghammer-stovring2011}.  Such an object
consists of a family $\{X_n\}_{n \in \mathbb{N}}$ of sets along with
restriction functions $r_n : X_{n+1} \to X_n$ for all $n$.  (in
category theoretic terminology, these are presheaves on the poset of
natural numbers.)  We think of a $\mathbb{N}$-indexed set as an
infinite sequence of increasingly precise approximations to the true
type being modeled.
%
Key to guarded domain theory is that there is an operator
$\triangleright$ on step-indexed sets called ``later''. In terms of
sequences of approximations, the later operator delays the
approximation by one step. Then the crucial axiom of guarded domain
theory is that any guarded domain equation $X \cong F(\triangleright
X)$ has a unique solution. This allows guarded domain theory to model
essentially \emph{any} recursive concept, with the caveat that the
recursion is \emph{guarded} by a use of the later operator.

% The theorems we want are about the whole sequence of approximations
% whereas working with presheaves is about each finite approximation
% The statement of graduality is in terms of global elements
\eric{This next paragraph could be moved to the section on adequacy, but since
the example is in the analytic setting it also makes sense to keep it here. I'm
not sure which option makes more sense.} While the definitions in guarded domain
theory are constructed as sequences of approximations, results about semantics
of programs, e.g., graduality, are actually statements concerning entire
\emph{sequences} of approximations. As a simple example, consider the set of
programs that may take a computational step or return unit.\footnote{This
example is adapted from \cite{mogelberg-paviotti2016}.} In the topos of trees,
we can model the set of these programs as the indexed family of sets $\{X_n\}_{n
\in \mathbb{N}}$ where for each $n \ge 1$ we define $X_n = \{0,1,\dots,n-1,
\bot\}$. Here, $i$ denotes a program that steps $i$ times and then returns, and
$\bot$ denotes a program that fails to terminate in $n-1$ steps or fewer. For
any fixed $n$, the set $X_n$ fails to distinguish a program that takes $n$ steps
and then terminates from a program that never terminates. On the other hand, if
we define the denotation of a program to be a \emph{sequence} of elements $x_i
\in X_i$ for all $i \in \mathbb{N}$, then the denotation of the diverging

%% \max{what is the point you are trying to convey in this paragraph?}
%% Because the solutions we obtain working in guarded domain theory are constructed
%% as a sequence of increasingly better approximations, when we want to establish a
%% property that holds in the limit, we must reason about all finite
%% approximations. In the setting of the topos of trees, this manifests as the need
%% to consider \emph{global elements} of the presheaves, i.e., a family of elements
%% $x_i \in X_i$ compatible with the restriction maps. For example, consider the
%% set of programs that may take a computational step or return unit.\footnote{This example is adapted from \cite{mogelberg-paviotti2016}.} In the
%% topos of trees, we can model this set as the indexed family of sets $\{X_n\}_{n \in \mathbb{N}}$
%% where for each $n \ge 1$ we define $X_n = \{0,1,\dots,n-1, \bot\}$. Here, $i$ denotes a program that steps
%% $i$ times and then returns, and $\bot$ denotes a program that fails to terminate
%% in $n-1$ steps or fewer. For any fixed $n$, the set $X_n$ fails to distinguish a
%% program that takes $n$ steps and then terminates from a program that never
%% terminates. On the other hand, if we instead consider the denotation of a
%% program to be a global element of the presheaf $\{X_n\}_{n \in \mathbb{N}}$ -- a family of elements
%% $x_i \in X_i$ for all $i \in \mathbb{N}$ -- then the denotation of the diverging
%% >>>>>>> Stashed changes
program will be distinct from that of a program that terminates, regardless of
the number of steps it takes.
